Racets: Embedding Policy{-}Agnostic Programming in Racket

### 2018-07-20 paper/paper.tex
Adding formalization section math, mostly done now, but no prose, going to work on section 4 tonight
hacking on section 4

\usepackage{booktabs} % For formal tables
\usepackage[linguistics]{forest}
\usepackage{tikz}
\usetikzlibrary{cd}
\usepackage{xcolor}
\usepackage{mathpartir} % Inference rules
\usepackage[ruled]{algorithm2e} % For algorithms
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

@@

\newcommand{\comm}[3][\color{red}]{{#1{[{#2}: {#3}]}}}
\newcommand{\zw}[1]{\comm[\color{green!66!black}]{Zhanpeng}{#1}}
\newcommand{\kris}[1]{\comm[\color{blue}]{Kris}{#1}}
\newcommand{\tom}[1]{\comm[\color{orange}]{Tom}{#1}}
\newcommand{\lang}[0]{$\lambda_\textsc{FE}$\xspace}
\newcommand{\facet}[3]{{\fcol{\ensuremath{\langle \, {#1} \: ? \: {#2} \diamond \: {#3} \, \rangle}}}}
\newcommand{\formfacet}[3]{{\fcol{\ensuremath{\langle \langle \, {#1} \: ? \: {#2} \diamond \: {#3} \,  \rangle \rangle}}}}

@@

            \blue{fv} & \blue{\in} & \blue{\text{facet-val}} & \blue{::=} & %% \blue{bv \mid \facet{\alpha}{fv}{fv}}

@@

    \hfill \boxed{\blue{e, \rho, \sigma \Downarrow_{pc}^E \sigma, v}}

@@

@@

@@

                     v' = \fcol{read(\sigma',v,pc)}}

@@

                    {e_1, \rho, \sigma \Downarrow_{pc}^E v_1, \sigma' \\\\

@@

                     \sigma''' = \fcol{write(\sigma'', v_1, pc, v_2)}}

@@

@@

          \inferrule[\textsc{Let-Label}]
          {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', v \\\\
           \alpha \not\in dom(\sigma) \and \sigma' = \sigma [\alpha \mapsto v] \\\\
           e_2, \rho [ \ell \mapsto \alpha ], \sigma' \Downarrow_{pc}^E \sigma'', v'}
          {(\code{let-label} ~ \ell ~ e_1 ~ e_2), \rho, \sigma \Downarrow_{pc}^E \sigma'', v'}}

@@

          \inferrule[\textsc{Fac-Create}]
          {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', l \\\\
           e_3, \rho, \sigma'' \Downarrow_{pc \cup \{-l\}}^E \sigma''', v_2 }
          {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma''', \formfacet{pc}{v_1}{v_2}}}

@@

          \inferrule[\textsc{Obs}]
          {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \\\\
           e_2, \rho, \sigma' \Downarrow_{pc}^E \sigma'', v  \\\\
           \langle (\lambda x.~e), \rho' \rangle = \sigma''(\ell) \\\\
           e_3, \rho' [x \mapsto v], \sigma'' \Downarrow_{pc}^E \sigma''', v'}
          {(\code{obs} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma''' v'}}
              %%\inferrule[]
\end{figure*}

>>>>

\usepackage{booktabs} % For formal tables
\usepackage[linguistics]{forest}
\usepackage{tikz}
\usetikzlibrary{cd}
\usepackage{xcolor}
\usepackage{mathpartir} % Inference rules
\usepackage[ruled]{algorithm2e} % For algorithms
\usepackage{setspace}
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

@@

\newcommand{\comm}[3][\color{red}]{{#1{[{#2}: {#3}]}}}
\newcommand{\zw}[1]{\comm[\color{green!66!black}]{Zhanpeng}{#1}}
\newcommand{\kris}[1]{\comm[\color{blue}]{Kris}{#1}}
\newcommand{\tom}[1]{\comm[\color{orange}]{Tom}{#1}}
\newcommand{\lang}[0]{$\lambda_\textsc{FE}$\xspace}
\newcommand{\facet}[3]{{\fcol{\ensuremath{\langle \, {#1} \: ? \: {#2} \diamond \: {#3} \, \rangle}}}}
\newcommand{\formfacet}[3]{{\fcol{\ensuremath{\langle \langle \, {#1} \: ? \: {#2} \diamond \: {#3} \,  \rangle \rangle}}}}
\newcommand{\var}[1]{\mathit{#1}}

@@

            \fcol{b} & \fcol{\in} & \fcol{\text{branch}} & \fcol{::=} & \fcol{+ \ell \mid - \ell} \\
            \fcol{pc} & \fcol{\in} & \fcol{\text{PC}} & \fcol{=} & \fcol{\wp(\text{branch})} \\
            \blue{\rho} & \blue{\in} & \blue{\text{env}} & \blue{=} & \blue{\text{var} \rightharpoonup v}\\
            \blue{\sigma} & \blue{\in} & \blue{\text{store}} & \blue{=} & \blue{\text{addr} \rightharpoonup v}

@@

    \hfill \textit{(Expression Evaluation)} \boxed{\blue{e, \rho, \sigma \Downarrow_{pc}^E \sigma, v}}

@@

          \inferrule[\textsc{Apply}]{e_1, \rho, \sigma \Downarrow_{pc}^E v_1, \sigma' \and
                                     \fcol{(v_1 v_2), \rho, \sigma'' \Downarrow_{pc}^A \sigma''', v'}}
                                    {(e_1 ~ e_2), \rho, \sigma \Downarrow_{pc}^E \sigma''', v'}

@@

                    {e, \rho, \sigma \Downarrow_{pc}^E v, \sigma' \and

@@

                     v' = \fcol{\var{read}(\sigma',v,pc)}}

@@

                    {e_1, \rho, \sigma \Downarrow_{pc}^E v_1, \sigma' \and

@@

                     \sigma''' = \fcol{\var{write}(\sigma'', v_1, pc, v_2)}}

@@

       \hfill \textit{(Facet Creation)}

@@

          \inferrule[\textsc{Fac-Create-Split}]
          {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \and
            \{+\ell, -\ell\} \cup pc = \varnothing \\\\
           e_3, \rho, \sigma'' \Downarrow_{pc \cup \{-l\}}^E \sigma''', v_2 \and
           v = \formfacet{pc}{v_1}{v_2}}
          {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma''', v}}

@@

          \inferrule[\textsc{Fac-Create-Left}]
          {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \and
           + \ell \in pc \\\\
           e_2, \rho, \sigma \Downarrow_{pc}^E \sigma'', v}
          {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma'', v}}

@@

          \inferrule[\textsc{Fac-Create-Right}]
          {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \and
           - \ell \in pc \\\\
           e_3, \rho, \sigma \Downarrow_{pc}^E \sigma'', v}
          {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma'', v}}     \end{array}

      \end{array}
    \end{displaymath}
    \\
    \hfill \textit{(Label Creation / Observation)}
    \\
    \[
    \begin{array}{cc}
      {\color{\colorFAC}
        \inferrule[\textsc{Let-Label}]
          {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \langle \lambda x. ~ e, \rho' \rangle \\\\
           \alpha \not\in dom(\sigma') \and \sigma'' = \sigma' [\alpha \mapsto \langle \lambda x. ~ e, \rho' \rangle] \\\\
           e_2, \rho [ \ell \mapsto \alpha ], \sigma'' \Downarrow_{pc}^E \sigma''', v}
          {(\code{let-label} ~ \ell ~ e_1 ~ e_2), \rho, \sigma \Downarrow_{pc}^E \sigma''', v}}
          & 
          {\color{\colorFAC}
          \inferrule[\textsc{Obs}]
                    {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell
                      \and e_2, \rho, \sigma' \Downarrow_{pc}^E \sigma'', v  \\\\
                      \langle (\lambda x.~e), \rho' \rangle = \sigma''(\ell) \\\\
                      e, \rho' [x \mapsto v], \sigma'' \Downarrow_{pc}^E \sigma''', v^{\pm} \\\\
                      e_3, \rho, \sigma''' \Downarrow_{pc}^E \sigma'''', v' \\\\
                      v'' = \fcol{\var{obs}(\ell, pc, v', v^{\pm})}}
                    {(\code{obs} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma'''', v''}}
    \end{array}
    \]
    \\
    \hfill \textit{(Possibly-Faceted Application)} \boxed{\fcol{(v ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma, v}}
    \\
    \begingroup
    \color{\colorMATH}
    \begin{displaymath}
      \begin{array}{c}
        \begin{array}{ccc}
          {\color{\colorMATH}
          \inferrule[\textsc{App-$\star$}]
          { }
          {(\star ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma, \star}}
          & 
          {\color{\colorMATH}
          \inferrule[\textsc{App-Base}]
          {e, \rho'[x \mapsto v], \sigma \Downarrow_{pc}^E \sigma', v' }
          {(\langle \lambda x.~e, \rho' \rangle ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}}
          & 
          {\color{\colorFAC}
          \inferrule[\textsc{App-Split}]
          {\{+\ell, -\ell\} \cap pc = \varnothing \and
            (v^+ ~ v), \rho, \sigma \Downarrow_{pc \cup \{+\ell\}}^A \sigma', {v^+}' \\\\
            (v^- ~ v), \rho, \sigma' \Downarrow_{pc \cup \{-\ell\}}^A \sigma'', {v^-}' \and
            v' = \formfacet{\{+ \ell\}}{{v^+}'}{{v^-}'}}
          {(\facet{\ell}{v^+}{v^-} ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma'', v'}}
        \end{array}
        \\
        \begin{array}{cc}
          {\color{\colorFAC}
          \inferrule[\textsc{App-Facet-Pos}]
          {+\ell \in pc \\\\
            (v^+ ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}
          {(\facet{\ell}{v^+}{v^-} ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}}
          & 
          {\color{\colorFAC}
          \inferrule[\textsc{App-Facet-Neg}]
          {-\ell \in pc \\\\
            (v^- ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}
          {(\facet{\ell}{v^+}{v^-} ~ v), \rho, \sigma \Downarrow_{pc}^A \sigma', v'}}
        \end{array}
      \end{array}
    \end{displaymath}
              %%\inferrule[]
    \endgroup
    \label{fig:semantics}
\end{figure*}

\section{Faceted Execution as Macros}

The semantics of faceted execution is an extension of the lambda
calculus, leading to a natural question: can we use Racket's macros to
extend Racket to faceted execution? We will see that the answer is
yes, and the translation from the big-step rules is surprisingly
straightforward. This section of our paper describes the design of
Racets, a prototype implementation of faceted execution using Racket
macros. In Section~\ref{sec:conclusion} we remark upon current
directions scaling Racets to the whole of Racket.

\paragraph*{Choosing a Representation for Facets, Labels, and Program Counters}

In setting out to implement facets, we must first choose how we will
represent facets, labels, and program counters. We have chosen to
implement facets simply as Racket \lstinline|struct|s, containing a
label along with positive and negative branches:

\begin{lstlisting}[language=Racket,escapechar=|,name=racets]
(struct facet (labelname left right))
\end{lstlisting}

Next, we must choose a representation of labels. At first
consideration, it appears sensible to represent labels simply as
closures. After all, labels are simply used as predicates testing
whether or not to reveal a facet's positive or negative
branch. However, there is a subtle issue: facet canonicalization
relies upon a total order for labels. In light of this, we represent
labels as a pair of a name and a policy:

\begin{lstlisting}[language=Racket,escapechar=|,name=racets]
(struct labelpair (name pol))
\end{lstlisting}

Now that we have defined labels, we can define branches, which are
positive or negative labels:

\begin{lstlisting}[language=Racket,escapechar=|,name=racets]
(struct pos (lab))
(struct neg (lab))
\end{lstlisting}

Similarly, program counters are sets of branches. However, we must
still ask how we will keep track of the ``current'' program
counter. Our implementation uses Racket's parameters, though other
mechanisms (such as continuation marks\kris{cite}, to which parameters
macro-expand) could also be used. Racets defines the parameter
\code{current-pc}, and updates it as computation progresses:

\begin{lstlisting}[language=Racket,escapechar=|,name=racets]
(define current-pc (make-parameter (set)))
\end{lstlisting}

\paragraph*{Facet Creation}

Facet creation appears as three separate rules in
Figure~\ref{fig:semantics}, which we recapitulate here in three
distint colors for each case:

\[
\begin{array}{ccc}
  {\color{brown}
    \inferrule[\textsc{Fac-Create-Left}]
              {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \and
                \boxed{+ \ell \in pc} \\\\
                e_2, \rho, \sigma \Downarrow_{pc}^E \sigma'', v}
              {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma'', v}}
  & 
  {\color{gray}
    \inferrule[\textsc{Fac-Create-Right}]
              {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \and
                \boxed{- \ell \in pc} \\\\
                e_3, \rho, \sigma \Downarrow_{pc}^E \sigma'', v}
              {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma'', v}}
  &
  {\color{purple}
    \inferrule[\textsc{Fac-Create-Split}]
              {e_1, \rho, \sigma \Downarrow_{pc}^E \sigma', \ell \and
                \boxed{\{+\ell, -\ell\} \cup pc = \varnothing} \\\\
                e_2, \rho, \sigma' \Downarrow_{pc \cup \{+l\}}^E \sigma'', v_1 \\\\
                e_3, \rho, \sigma'' \Downarrow_{pc \cup \{-l\}}^E \sigma''', v_2 \and
                v = \formfacet{pc}{v_1}{v_2}}
              {(\code{fac} ~ e_1 ~ e_2 ~ e_3), \rho, \sigma \Downarrow_{pc}^E \sigma''', v}}

\end{array}
\]

Translating these rules to Racket involves observing that each of
these rules will apply under one of three disjoint circumstances (each
of them boxed in the above rules): $+\ell \in pc$, $-\ell \in pc$, or
else $\{+\ell,-\ell\} \cap pc = \varnothing$. This leads to a
relatively straightforward implementation that inspects the current
program counter to decide which rule to implement:

\begin{lstlisting}[language=Racket,escapechar=|,name=racets]
(define-syntax-rule (fac l e1 e2)
  (cond
|{\color{brown}|    [(set-member? (current-pc) (pos (facet-labelname l))) e1] |}|
|{\color{gray}|    [(set-member? (current-pc) (neg (facet-labelname l))) e2]  |}|

|{\color{purple}|    [else
     (let ([left
            (parameterize ([current-pc (set-add (current-pc)
                                                (pos (facet-labelname l)))]))]
           [right
            (parameterize ([current-pc (set-add (current-pc)
                                                (neg (facet-labelname l)))]))]
           (mkfacet (labelpair-name l) v1 v2)))])) |}|
\end{lstlisting}

As our definition is a transliteration of that in Figure 6 of Austin
et. al's paper~\cite{Austin:2012}, we omit the implementation here.
The result is a function, \code{mkfacet}, which (given a label and
both positive and negative branches) forms a facet:

\begin{lstlisting}[language=Racket,escapechar=|,name=racets]
(define (mkfacet label v1 v2)
  (canonicalize (set-add (current-pc) (pos name)) v1 v2))
\end{lstlisting}
